---
title: "Interactive: U.S. Suicide trends over time"
author: "R. Chris Berg"
output: 
  html_document: 
    theme: flatly
---

```{r setup, include=FALSE}

library(tidyverse)
library(janitor)
library(ggplot2)
library(plotly)
library(hrbrthemes)
library(sf)
library(albersusa)
library(scales)
library(htmltools)

suicide_data <- 
  read_csv("D:/Economics/Projects/medicaid/data/raw/cdc_wonder_suicide_raw.csv") %>%
  clean_names() %>%
  filter( is.na(notes) == F & is.na(year)==F & state_code != 11 ) %>%
  select(-c("cause_of_death",
            "cause_of_death_code",
            "notes")) %>%
  mutate( across(-state, as.numeric))

state_suicide_2000 <- 
  suicide_data %>%
  filter( year == 2000)%>%
  select( state, age_adjusted_rate ) %>%
  rename("name"="state",
         "adj_rate_2000"="age_adjusted_rate")

state_suicide_2018 <- 
  suicide_data %>%
  filter( year == 2018)%>%
  select( state, age_adjusted_rate ) %>%
  rename("name"="state",
         "adj_rate_2018"="age_adjusted_rate")

state_suicide_growth_df <- 
  left_join(state_suicide_2000,state_suicide_2018) %>%
  mutate( suicide_rate_growth = round(100*(adj_rate_2018/adj_rate_2000-1),
                                      2)
          )
us_map_sf <- 
  st_as_sf(usa_composite(proj = "laea")) %>%
  select( name, geo_id, geometry ) %>%
  merge(state_suicide_growth_df)

```

*The data used here comes from the Centers for Disease Control [WONDER Database](https://wonder.cdc.gov/).*

## Suicide in the United States, 1999-2018

The United States has been experiencing a drastic increase in the rate of suicide, with over 1 in 10,000 US citizens dying by some form of suicide in 2018. 

```{r national_rate, echo=F, message=F, dpi=300}
suicide_data %>% 
  group_by(year) %>%
  summarise( pop = sum(population) ,
             suicides = sum(deaths) ) %>%
  mutate( crude_rate_total = (suicides/pop)*100000 
          ) %>%
  ggplot( aes( x = year , 
             y = crude_rate_total ) ) + 
  geom_line( size = 1 ,
          color = 'red') + 
  labs( title = "United States suicide rate over time",
        caption = "(Data from the CDC WONDER Database)",
        x = "Year" , 
        y = "Deaths per 100,000 population")+
  theme_ipsum_rc()
```

Despite [public concern](https://www.usatoday.com/story/news/nation/2020/01/30/u-s-suicide-rate-rose-again-2018-how-can-suicide-prevention-save-lives/4616479002/), a 2017 meta-analysis published in [the American Psychological Association's "Psychological Bulletin"](https://psycnet.apa.org/record/2016-54856-001) notes that "...predictive ability has not improved across 50 years of research;" which leaves researchers and practitioners with few tools to reverse course.

## Suicide rate among states, 1999-2018

Below is an interactive chart which shows the trend in the age-adjusted suicide rate for each state. You can hover-over the line for a particular state and a box will pop-up showing which state you're looking at, and what their age-adjusted suicide rate was for the given year.

To pull-up the line graph for a specific state, double-click the state name in the legend. Click (once) on any additional states to compare their trends.

```{r states_interactive, echo=FALSE, dpi=120}
state_rate_plot <- 
  ggplot( data = suicide_data , 
          aes(x = year , y = age_adjusted_rate , color = state )
          ) +
  geom_line() +
  labs( title = "Suicide rate by state",
        x = "Year" , 
        y = "Deaths per 100,000 population",
        color="State") +
  theme_ipsum_rc()

ggplotly(state_rate_plot)
```

## Geographic distribution of suicide rate trends

Which states have seen the largest increases since the year 2000? Below is an interactive map showing the percentage growth in the (age-adjusted) number of suicide deaths per 100,000 citizens of a state. Lighter colors indicate a higher percentage growth since 2000.

```{r map_interactive, echo=FALSE, warning=F, message=F,dpi=120}

state_suicide_growth_dist <- 
  ggplot( data = us_map_sf , 
          ) +
  geom_sf(aes( fill = suicide_rate_growth , label=name) , col="white") +
  scale_fill_viridis_c(name="Percent change <br> 2000-2018",
                       labels = scales::label_percent(scale=1))+
  ggtitle("Geography of suicide rate trends") +
  theme_ipsum_rc()

ggplotly(state_suicide_growth_dist)
```